Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-06-27 Manuscript/manuscript.tex
update
saving work

The statistical errors for the excitation energies, for the different CI models considered here,
are presented in Tab.~\ref{tab:1} and Tab.~\ref{tab:2},
%organized by their computational scaling,
for singly excited states of closed-shell and open-shell systems, respectively.
There are two key elements that explain the main trends on the observed statistical errors.
The first concerns the set of molecular orbitals.

hCI1 and hCI1.5 largely overestimate the excitation energies of closed-shell systems (see Tab.~\ref{tab:1}).
The excitations in these low-order hCI models already account for some correlation of the ground state, but represent orbital relaxation of the excited state, which in turn remains less correlated.
This favors the description of the ground state and explains the overestimated excitation energies, similarly to what is found for CISD. \cite{}
%As will be further discussed below, this effect parallels
A state-specific approach would be expected to eliminate this bias and reduce the errors, as observed, for instance, when going from CISD to $\Delta$CISD \cite{Kossoski_2023}.
However, we found that $\Delta$hCI1 and $\Delta$hCI1.5 produced even larger MAEs than with the ground-state based approach, though this time by largely underestimating the excitation energies.
This can be understood from the different classes of determinants accessed from the ground state reference (closed-shell determinant) and excited state reference (single open-shell CSF).
Lower orders of hCI therefore capture a larger fraction of correlation for the latter type of reference, and are thus not suitable for calculations of excitation energies for closed-shell systems.

Ramping up to $\Delta$hCI2, this inbalance is largely reduced, even though the excitation energies are still underestimated.
$\Delta$hCI2 and $\Delta$CISD display comparable MAEs, despite the somewhat smaller absolute MSE and RMSE of the latter.
This means that the additional determinants present in the former model recover a larger fraction of the correlation energy for excited state than for the ground state, thus creating a bias toward underestimated excitation energies.
%Even though this unbalance is still observed for $\Delta$hCI2, it is much less serious than observed for the lower-order state-specific hCI methods.

@@

\begin{table}[ht!]
\caption{Mean Signed Error (MSE), Mean Absolute Error (MAE), and Root-Mean Square Error (RMSE) in Units of eV, with Respect to Reference Theoretical Values, for the Set of 
38 Singly Excited States of Closed-Shell Systems and 
Listed in the {\SupInf}.
}
\label{tab:1}
\begin{ruledtabular}
\begin{tabular}{lddd}
method            & \mc{1}{c}{MSE} & \mc{1}{c}{MAE} & \mc{1}{c}{RMSE} \\
\hline
$\Delta$CSF       & -0.75 & 0.81 & 0.95 \\
CIS               & +0.03 & 0.62 & 0.63 \\
\hline
hCI1              & +1.08 & 1.12 & 1.25 \\
$\Delta$hCI1      & -1.37 & 1.41 & 1.52 \\
hCI1.5            & +1.95 & 1.95 & 2.03 \\
$\Delta$hCI1.5    & -2.95 & 2.95 & 3.14 \\
$\Delta$hCI2      & -0.19 & 0.20 & 0.25 \\
$\Delta$CISD      & -0.13 & 0.17 & 0.22 \\
\hline
sCI2/sCI2         & +1.40 & 1.40 & 1.53 \\
sCI2/sCI0         & -0.22 & 0.42 & 0.57 \\
$\Delta$sCI2/sCI2 & +0.74 & 0.77 & 0.85 \\
$\Delta$sCI2/sCI0 & -0.88 & 0.88 & 0.96 \\
\end{tabular}
\end{ruledtabular}
\end{table}

The statistical errors for the excitation energies of open-shell systems are shown in Tab.~\ref{tab:2}.

\begin{table}[ht!]
\caption{Mean Signed Error (MSE), Mean Absolute Error (MAE), and Root-Mean Square Error (RMSE) in Units of eV, with Respect to Reference Theoretical Values, for the Set of 
16 Singly-Excited States from Open-Shell Doublets 
Listed in the {\SupInf}.
}
\label{tab:2}
\begin{ruledtabular}
\begin{tabular}{lddd}
method            & \mc{1}{c}{MSE} & \mc{1}{c}{MAE} & \mc{1}{c}{RMSE} \\
\hline
$\Delta$CSF       & -0.09 & 0.50 & 0.67 \\
CIS               & +0.30 & 0.41 & 0.55 \\
\hline
hCI1              & +0.76 & 0.80 & 1.02 \\
$\Delta$hCI1      & -0.09 & 0.58 & 0.94 \\
hCI1.5            & +1.22 & 1.22 & 1.45 \\
$\Delta$hCI1.5    & -0.26 & 0.66 & 0.92 \\
$\Delta$hCI2      & -0.03 & 0.14 & 0.19 \\
$\Delta$CISD      & +0.00 & 0.16 & 0.23 \\
\hline
sCI1              & +0.27 & 0.44 & 0.66 \\
$\Delta$sCI1      & -0.15 & 0.28 & 0.40 \\
\end{tabular}
\end{ruledtabular}
\end{table}

Despite the rather poor performance of lower-order hCI models for excitation energies,
we further investigated whether higher-order versions could be competitive against the traditional excitation-based counterparts.
For that, we performed calculations for a subset of 14 singly-excited states of small closed-shell molecular systems (\ce{H2O}, \ce{BF}, \ce{HCF}, \ce{HCl}, and \ce{N2}),
with corresponding statistical errors shown in Tab.~\ref{tab:3}.
Even though this is a small set, it should be enough to obtain the main trends on the accuracy of higher-order models.

\begin{table}[ht!]
\caption{Mean Signed Error (MSE), Mean Absolute Error (MAE), and Root-Mean Square Error (RMSE) in Units of eV, with Respect to Reference Theoretical Values,
for the Set of 14 Singly Excited States of Closed-Shell Systems Listed in the {\SupInf}.
}
\label{tab:3}
\begin{ruledtabular}
\begin{tabular}{lddd}
method            &     \mc{1}{c}{MSE} & \mc{1}{c}{MAE} & \mc{1}{c}{RMSE} \\
\hline
CISD              & +4.09 & 4.09 & 4.18  \\
CISDT             & +0.09 & 0.18 & 0.20  \\
CISDTQ            & +0.14 & 0.14 & 0.16  \\
%
$\Delta$CISD      & -0.13 & 0.17 & 0.21  \\
$\Delta$CISDT     & -0.19 & 0.19 & 0.23  \\
$\Delta$CISDTQ    & -0.02 & 0.02 & 0.02  \\
%
%hCI1              &       &      &      \\
%hCI1.5            &       &      &      \\
hCI2              & +2.89 & 3.00 & 3.28  \\
hCI2.5            & +1.93 & 1.93 & 2.05  \\
hCI3              & +0.18 & 0.18 & 0.19  \\
%
%$\Delta$hCI1      &       &      &      \\
%$\Delta$hCI1.5    &       &      &      \\
$\Delta$hCI2      & -0.22 & 0.22 & 0.27  \\
$\Delta$hCI2.5    & -0.27 & 0.27 & 0.30  \\
$\Delta$hCI3      & -0.21 & 0.21 & 0.24  \\
$\Delta$hCI3.5    & -0.08 & 0.08 & 0.09  \\
\end{tabular}
\end{ruledtabular}
\end{table}

The case of $\Delta$hCI2 was discussed above for the larger set of excitation energies.
Here we further notice that hCI2 was found to be somewhat more accurate than CISD, even though the errors are still too large.
However, this improvement does not reflect on the state specific route (compare $\Delta$hCI2 and $\Delta$CISD in Tabs.~\ref{tab:1} and ~\ref{tab:2}).
Moving to the next order, even though the state-specific route ($\Delta$hCI2.5) is far more accurate than the ground-state based route (hCI2.5),
it is actually less accurate then $\Delta$hCI2.
An improved performance is found for hCI3, $\Delta$hCI3, CISDT, and $\Delta$CISDT, which present quite similar MAEs and RMSEs.
There is thus no advantage in moving from excitation-based to hCI at this order, similarly to what discussed before for the $\Delta$hCI2 and $\Delta$CISD models.
Moreover, $\Delta$hCI3 is not more accurate than hCI3, as also observed for $\Delta$CISDT and CISDT. \cite{Kossoski_2023}
We notice, however, that the state-specific models present negative MSEs, in contrast to the positives values obtained in the ground-state based route.
This reflects an umbalanced description of correlation effects, favoring the excited states.
At the next order, $\Delta$hCI3.5 has significantly smaller errors, but at a much more elevated computational cost.

>>>>

For each CI model considered here, we evaluated the mean signed error (MSE), mean absolute error (MAE), and root-mean square error (RMSE) associated with the excitation energy.
The statistical errors are presented in Tab.~\ref{tab:1} and Tab.~\ref{tab:2},
for singly excited states of closed-shell and open-shell systems, respectively.
For the lower order models, the calculations were performed for 46 (closed-shell) and 20 (open-shell) excited states.
In turn, a subset of 14 (closed-shell) and 5 (open-shell) excited states was considered for the higher order models, given their more intensive computational cost.
Even though these subsets are too small for meaningful absolute statistics, they can still show the main trends on the higher order CI models.
%A detailed comparison of excitation-based CI along the ground-state and the state-specific routes can be found elsewhere. \cite{Kossoski_2023}
%The focus of the present discussion lies on the comparison between hCI, sCI, and excitation-based CI, for both routes,
%
%There are two key elements that explain the main trends on the observed statistical errors. The first concerns the set of molecular orbitals.

We start the discussion with the excitations from closed-shell systems, with corresponding statistics shown in Tab.~\ref{tab:1}.
The first level of hCI, hCI1, produces poor excitation energies, with a MAE of \SI{1.07}{\eV}.
This model shares the same computational scaling with CIS, but includes the paired double excitations, 
which clearly do not improve the descent CIS results (MAE of \SI{0.61}{\eV}).
Moving one rank up, to hCI1.5, the MAE increases to \SI{1.91}{\eV}, and reaches a maxium of \SI{3.53}{\eV} at the hCI2 level.
From that point on, the errors systematically decrease.
hCI2.5 is stil as inaccurate (MAE of \SI{1.95}{\eV}) as hCI1.5.
Finally, at the hCI3 level, significantly smaller errors are achieved (MAE of \SI{0.19}{\eV}),
which are further reduced at the hCI3.5 level (MAE of \SI{0.11}{\eV}).

The errors of the hCI models thus increase as one augments the hierarchy parameter $h$, reaching a maximum at hCI2, and then decrease towards higher orders.
This behavior paralles what is found with excitation-based CI, where CISD is far less accurate than CIS, which in turn is less accurate than CISDT and CISDTQ.
This can be understood based on the role of the excited determinants for ground and excited states.
%
The excited determinants of the low-order hCI models (hCI1 and hCI1.5) already account for some correlation of the ground state, but represent mostly orbital relaxation of the excited state, which thus remains less correlated.
This favored description of the ground state explains the overestimated excitation energies.
The effect becomes exagerated at the hCI2 (and CISD) models, since it captures most of the ground state correlation through the unpaired double excitations.
Higher orders models are needed to recover most of the excited state correlation and the excitation energy errors thus decrease.
%(though always overestimating the reference values).
The same argument also explains why hCI systematically overestimates the excitation energies, just as excitation-based CI does.

The trends discussed above are directly linked with the use of ground-state HF orbitals.
Employing state-specific orbitals would be expected to suppress this bias and improve the accuracy, as observed, for instance, when going from CISD to $\Delta$CISD \cite{Kossoski_2023}.
For the lower orders of state-specific hCI ($\Delta$hCI1 and $\Delta$hCI1.5), we actually found larger errors than with the ground-state based approach, though by underestimating the excitation energies.
Besides the set of orbitals, a second key aspect must be considered.
For a given hierarchy parameter $h$, different classes of determinants are accessed from the ground state reference (a closed-shell determinant) and from the excited state reference (a single open-shell CSF).
The case of $\Delta$hCI1 is illustrated in Fig.~\ref{fig:DhCI1}.
%For the ground state calculation, it accounts for two classes of excited determinants, generated by single excitation and by paired double excitation from the closed-shell reference.
%For the singly-excited state, four types of determinants are generated. On top of the single and paired double excitations from the doubly-occupied orbitals, 
There is arguably more diversity in the types of determinants employed for the excited-state calculation than for the ground-state one.
Because of that, lower orders of hCI may be expected to capture a larger fraction of correlation for the former calculation.
%and are thus not suitable for calculations of excitation energies for closed-shell systems.
Although hCI is constructed to account for all classes of determinants whose number share the same computational scaling,
at the lower orders this procedure does not lead to a balanced description of correlation, at least for excitations of closed-shell systems.
As will be discussed later for the excitations of open-shell systems, this inbalance probably arises from the distinct reference determinants employed for the closed-shell excitations.
%
Another important point to realize is that a state-specific approach is not necessarily more accurate than a ground-state based approach.
Both the orbitals and the classes of determinants of a given CI model play an important role in the fine balance of ground- and excited-state correlation effects.

Ramping up to $\Delta$hCI2, this inbalance is reduced to a great extent, even though the excitation energies are still underestimated.
%Even though this unbalance is still observed for $\Delta$hCI2, it is much less serious than observed for the lower-order state-specific hCI methods.
$\Delta$hCI2 display a MAE of \SI{0.20}{\eV}, comparable to that of $\Delta$CISD (\SI{0.17}{\eV}) display comparable MAEs.
Albeit small, this difference is probably statistically significant, and $\Delta$CISD should still outperform $\Delta$hCI2 by some margin.
Similarly, $\Delta$hCI2 has a more negative MSE (\SI{-0.19}{\eV} than $\Delta$CISD (\SI{-0.13}{\eV})
%This means that the additional determinants present in the former model recover a larger fraction of the correlation energy for excited state than for the ground state, thus creating a bias toward underestimated excitation energies.
Ramping up to $\Delta$hCI2.5 increases the errors (MAE of \SI{0.27}{\eV}),
which is probably due to another layer of inbalance concerning the (ground-state and excited-state) determinants that enter at this stage.
(Despite the different number of states considered, the observed difference between $\Delta$hCI2 and $\Delta$hCI2.5 is statistically significant.)
The situation improves at $\Delta$hCI3 (MAE of \SI{0.21}{\eV}),
while $\Delta$hCI3.5 generates considerably more accurate results (MAE of \SI{0.08}{\eV}),

%Despite the rather poor performance of lower-order hCI models for excitation energies,
%we further investigated whether higher-order versions could be competitive against the traditional excitation-based counterparts.
%For that, we performed calculations for a subset of 14 singly-excited states of small closed-shell molecular systems (indicated in the {\SupInf}).
%(\ce{H2O}, \ce{BF}, \ce{HCF}, \ce{HCl}, and \ce{N2}),
%with corresponding statistical errors shown in Tab.~\ref{tab:1}.
%Even though this is a small set, it should be enough to obtain the main trends on the accuracy of higher-order models.

\begin{table}[ht!]
\caption{Mean Signed Error (MSE), Mean Absolute Error (MAE), and Root-Mean Square Error (RMSE) in Units of eV, with Respect to Reference Theoretical Values, for the Set of 
Singly Excited States of Closed-Shell Systems Listed in the {\SupInf}.
}
\label{tab:1}
\begin{ruledtabular}
\begin{tabular}{ldddd}
method            & \mc{1}{c}{count} & \mc{1}{c}{MSE} & \mc{1}{c}{MAE} & \mc{1}{c}{RMSE} \\
\hline
CIS               & 46  & +0.04 & 0.61 & 0.60 \\
CISD              & 14  & +4.09 & 4.09 & 4.18 \\
CISDT             & 14  & +0.10 & 0.19 & 0.20 \\
CISDTQ            & 14  & +0.15 & 0.15 & 0.17 \\
%
\hline
hCI1              & 46  & +0.97 & 1.07 & 1.20 \\
hCI1.5            & 46  & +1.91 & 1.91 & 2.00 \\
hCI2              & 14  & +2.99 & 3.53 & 3.61 \\
hCI2.5            & 14  & +1.95 & 1.95 & 2.06 \\
hCI3              & 14  & +0.19 & 0.19 & 0.21 \\
hCI3.5            & 14  & +0.11 & 0.11 & 0.13 \\
%
\hline
$\Delta$CSF       & 46  & -0.73 & 0.80 & 0.94 \\
$\Delta$CISD      & 46  & -0.13 & 0.17 & 0.22 \\
$\Delta$CISDT     & 14  & -0.20 & 0.20 & 0.22 \\
$\Delta$CISDTQ    & 14  & -0.02 & 0.02 & 0.02 \\
%
\hline
$\Delta$hCI1      & 46  & -1.36 & 1.40 & 1.50 \\
$\Delta$hCI1.5    & 46  & -2.91 & 2.91 & 3.10 \\
$\Delta$hCI2      & 46  & -0.19 & 0.20 & 0.25 \\
$\Delta$hCI2.5    & 14  & -0.27 & 0.27 & 0.30 \\
$\Delta$hCI3      & 14  & -0.21 & 0.21 & 0.24 \\
$\Delta$hCI3.5    & 14  & -0.08 & 0.08 & 0.09 \\
%
\hline
sCI2/sCI2         & 46  & +1.26 & 1.34 & 1.49 \\
sCI2/sCI0         & 46  & -0.41 & 0.57 & 0.89 \\
$\Delta$sCI2/sCI2 & 46  & +0.69 & 0.79 & 0.90 \\
$\Delta$sCI2/sCI0 & 46  & -0.99 & 0.99 & 1.15 \\
%
\end{tabular}
\end{ruledtabular}
\end{table}

We now move to the discussion about the excitation energies of open-shell systems.
The statistical errors are shown in Tab.~\ref{tab:2}.
%
The case of $\Delta$hCI2 was discussed above for the larger set of excitation energies.
Here we further notice that hCI2 was found to be somewhat more accurate than CISD, even though the errors are still too large.
However, this improvement does not reflect on the state specific route (compare $\Delta$hCI2 and $\Delta$CISD in Tabs.~\ref{tab:1} and ~\ref{tab:2}).
Moving to the next order, even though the state-specific route ($\Delta$hCI2.5) is far more accurate than the ground-state based route (hCI2.5),
it is actually less accurate then $\Delta$hCI2.
An improved performance is found for hCI3, $\Delta$hCI3, CISDT, and $\Delta$CISDT, which present quite similar MAEs and RMSEs.
There is thus no advantage in moving from excitation-based to hCI at this order, similarly to what discussed before for the $\Delta$hCI2 and $\Delta$CISD models.
Moreover, $\Delta$hCI3 is not more accurate than hCI3, as also observed for $\Delta$CISDT and CISDT. \cite{Kossoski_2023}
We notice, however, that the state-specific models present negative MSEs, in contrast to the positives values obtained in the ground-state based route.
This reflects an umbalanced description of correlation effects, favoring the excited states.
At the next order, $\Delta$hCI3.5 has significantly smaller errors, but at a much more elevated computational cost.

\begin{table}[ht!]
\caption{Mean Signed Error (MSE), Mean Absolute Error (MAE), and Root-Mean Square Error (RMSE) in Units of eV, with Respect to Reference Theoretical Values, for the Set of 
Singly-Excited States from Open-Shell Doublets Listed in the {\SupInf}.
}
\label{tab:2}
\begin{ruledtabular}
\begin{tabular}{ldddd}
method            & \mc{1}{c}{count} & \mc{1}{c}{MSE} & \mc{1}{c}{MAE} & \mc{1}{c}{RMSE} \\
\hline
CIS               & 20 & +0.33 & 0.36 & 0.50 \\
CISD              & 5  & +1.98 & 1.98 & 2.16 \\
CISDT             & 5  & +0.02 & 0.03 & 0.05 \\
CISDTQ            & 5  & +0.03 & 0.03 & 0.05 \\
%
\hline
hCI1              & 20 & +0.70 & 0.77 & 0.96 \\
hCI1.5            & 20 & +1.11 & 1.11 & 1.30 \\
hCI2              & 5  & +0.69 & 0.69 & 0.85 \\
hCI2.5            & 5  & +0.01 & 0.03 & 0.04 \\
hCI3              & 5  & -0.00 & 0.02 & 0.03 \\
hCI3.5            & 5  & -0.00 & 0.01 & 0.02 \\
hCI4              & 5  &       &      &      \\
%
\hline
$\Delta$CSF       & 20 & -0.04 & 0.41 & 0.58 \\
$\Delta$CISD      & 20 & +0.01 & 0.12 & 0.21 \\
$\Delta$CISDT     & 5  & -0.05 & 0.05 & 0.09 \\
$\Delta$CISDTQ    & 5  & -0.02 & 0.02 & 0.02 \\
%
\hline
$\Delta$hCI1      & 20 & -0.19 & 0.34 & 0.48 \\
$\Delta$hCI1.5    & 20 & -0.10 & 0.46 & 0.70 \\
$\Delta$hCI2      & 20 & -0.01 & 0.11 & 0.16 \\
$\Delta$hCI2.5    & 5  & -0.06 & 0.06 & 0.12 \\
$\Delta$hCI3      & 5  & -0.03 & 0.03 & 0.06 \\
$\Delta$hCI3.5    & 5  & -0.02 & 0.02 & 0.02 \\
$\Delta$hCI4      & 5  &       &      &      \\
%
\hline
sCI1              & 20 & +0.24 & 0.39 & 0.61 \\
$\Delta$sCI1      & 20 & -0.11 & 0.27 & 0.38 \\
\end{tabular}
\end{ruledtabular}
\end{table}

%\begin{table}[ht!]
%\caption{Mean Signed Error (MSE), Mean Absolute Error (MAE), and Root-Mean Square Error (RMSE) in Units of eV, with Respect to Reference Theoretical Values,
%for the Set of 14 Singly Excited States of Closed-Shell Systems Listed in the {\SupInf}.
%}
%\label{tab:3}
%\begin{ruledtabular}
%\begin{tabular}{lddd}
%method            &     \mc{1}{c}{MSE} & \mc{1}{c}{MAE} & \mc{1}{c}{RMSE} \\
%\hline
%CISD              & +4.09 & 4.09 & 4.18  \\
%CISDT             & +0.09 & 0.18 & 0.20  \\
%CISDTQ            & +0.14 & 0.14 & 0.16  \\
%$\Delta$CISD      & -0.13 & 0.17 & 0.21  \\
%$\Delta$CISDT     & -0.19 & 0.19 & 0.23  \\
%$\Delta$CISDTQ    & -0.02 & 0.02 & 0.02  \\
%\hline
%hCI1              & +1.62 & 1.62 & 1.65 \\
%hCI1.5            & +2.10 & 2.10 & 2.11 \\
%hCI2              & +2.89 & 3.00 & 3.28  \\
%hCI2.5            & +1.93 & 1.93 & 2.05  \\
%hCI3              & +0.18 & 0.18 & 0.19  \\
%$\Delta$hCI1      & -1.37 & 1.50 & 1.66 \\
%$\Delta$hCI1.5    & -2.62 & 2.62 & 2.72 \\
%$\Delta$hCI2      & -0.22 & 0.22 & 0.27  \\
%$\Delta$hCI2.5    & -0.27 & 0.27 & 0.30  \\
%$\Delta$hCI3      & -0.21 & 0.21 & 0.24  \\
%$\Delta$hCI3.5    & -0.08 & 0.08 & 0.09  \\
%\end{tabular}
%\end{ruledtabular}
%\end{table}

@@
